Dendrochronological dates confirm a Late Prehistoric population decline in the American Southwest derived from radiocarbon dates

### 2020-04-17 analysis/robinson2020.tex
using pander for colophon

% \usepackage{graphicx}

% \usepackage{fontspec,xunicode}

% \usepackage{polyglossia}
% \setdefaultlanguage[variant=usmax]{english}

@@

\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newenvironment{cslreferences}%
  {}%
  {\par}

@@

The spatial and temporal ubiquity of archaeological radiocarbon dates makes them an unprecedented proxy for human paleodemography. The past decade has witnessed a rapid increase in the use of archaeological radiocarbon time series to reconstruct prehistoric population growth and decline in various regions of the world {[}\protect\hyperlink{ref-Bevan2019}{1}--\protect\hyperlink{ref-Williams2012}{9}{]}. One of the first aggregations of radiocarbon dates for palaeodemographic analysis was by Berry {[}\protect\hyperlink{ref-Berry1982}{10}{]} in his study of population growth, decline and migration across the American Southwest. Radiocarbon time series were used as a complement to earlier periods that lacked robust dendrochronological records. Berry {[}\protect\hyperlink{ref-Berry1982}{10}{]} noted that they were ``not even remotely related to population size'', but rather the ``relative probability of occupation through time'' and ``the major trends in these charts (i.e., peaks and troughs) should provide a reliable indication of the direction of change''. Later, Rick {[}\protect\hyperlink{ref-Rick1987}{11}{]} more formally outlined the use of radiocarbon time series as a method for paleodemography by developing an `inference chain' that detailed the various biases involved in their use and interpretation. Rick identified three biases {[}\protect\hyperlink{ref-Rick1987}{11}{]}: creation bias, preservation bias, and investigation bias. Over the past decade, as archaeologists have assembled and analysed large radiocarbon datasets from various regions of the world, there has been considerable focus on these biases, notably creation {[}\protect\hyperlink{ref-Crombe2014}{12}--\protect\hyperlink{ref-Naudinot2014}{14}{]} and preservation bias {[}\protect\hyperlink{ref-Peros2010}{5},\protect\hyperlink{ref-Williams2012}{9},\protect\hyperlink{ref-Bluhm2019}{15}--\protect\hyperlink{ref-Surovell2009}{17}{]}, as well as bias caused by the radiocarbon calibration process {[}\protect\hyperlink{ref-Williams2012}{9},\protect\hyperlink{ref-Armit2013}{18}--\protect\hyperlink{ref-Timpson2014}{22}{]}. Of the three biases Rick identified, investigation bias has been the most understudied thus far. Investigation bias occurs when investigators gather more radiocarbon samples for particular periods of time or specific geographic regions, which leads to periods or regions with more or less radiocarbon representation than the average time period or region. The biased selection of radiocarbon samples would then appear to be relative increases or decreases in population when they are merely artefacts of research history and fieldwork practice. Investigation bias has been understudied because for many regions of the world radiocarbon data are the only currently available proxy evidence enabling reconstructions of paleodemography in continuous time series spanning millennia. Assessing the impact of investigation bias on radiocarbon-based reconstructions of paleodemography requires comparison with other paleodemographic proxies.

The American Southwest provides one the best available case studies for assessing the role of investigation bias in radiocarbon-based reconstructions of paleodemography. It also has one of the most well-documented cases of population growth and decline in the world {[}\protect\hyperlink{ref-Bocinsky2016}{23}--\protect\hyperlink{ref-Freeman2018PNAS}{27}{]}. For almost a century, researchers have collected multiple proxies, such as dendrochronological records, settlement size records, and ceramic seriation records to address the magnitude of this population growth and decline. However, a critical knowledge gap still exists for the exact scale of population decline across and beyond the Southwest {[}\protect\hyperlink{ref-Cameron2010}{28},\protect\hyperlink{ref-Boyer2010}{29}{]}. This gap is because the less intensively researched regions lack multiple proxies for robust reconstructions of paleodemography. In these regions, radiocarbon data are the only available proxy time series for reconstructing paleodemography. Over the past six years, we have collected all available radiocarbon data across and beyond the Southwest {[}\protect\hyperlink{ref-Robinson2019}{26}{]}. Initial analyses of this new radiocarbon dataset have revealed extensive population decline from AD 1300--1600 across and beyond the Southwest {[}\protect\hyperlink{ref-Robinson2019}{26},\protect\hyperlink{ref-Freeman2018PNAS}{27}{]}. Therefore, the geographic extent of Late Prehistoric population decline could be much broader than previously expected.

The potential importance of this widespread population decline for research on Late Prehistoric archaeology of the American West requires comparison of the sensitivity of lower-resolution radiocarbon data against other higher-resolution palaeodemographic proxies. In the most well-documented regions (such as the northern American Southwest) we expect radiocarbon records to reflect the highest amount of investigation bias, as researchers have the option of selecting from a suite of other higher-resolution proxies, including dendrochronology and ceramic seriation. Investigation bias can create an artificial `edge effect' spuriously appearing as a decrease in population. Here we develop a new method for arraying dendrochronological records as time series for comparison against radiocarbon time series. We focus on the broader region of the Upland US Southwest (UUSW) for which there is a robust dendrochronological record {[}\protect\hyperlink{ref-Bocinsky2016}{23}{]}.

We also focus on the more concentrated and intensively investigated Central Mesa Verde (CMV) and Northern Rio Grande (NRG) regions of the UUSW (Figure \ref{fig:Figure1}), for which multiproxy palaeodemographic reconstructions were produced by the Village Ecodynamics Project (VEP) {[}\protect\hyperlink{ref-Ortman2016}{30},\protect\hyperlink{ref-Schwindt2016}{31}{]}. The VEP is an interdisciplinary, multi-method research initiative aimed at better understanding demography of ancestral Pueblo farmers and human-environment interaction across the UUSW. As part of the project, VEP researchers cataloged data on all known archaeological habitation sites within a 4,600-km2 area in the CMV {[}\protect\hyperlink{ref-Schwindt2016}{31}{]} and a 6,900-km2 area in the NRG {[}\protect\hyperlink{ref-Ortman2016}{30}{]}, to which most of the CMV population migrated during the thirteenth century AD {[}\protect\hyperlink{ref-Ortman2016}{30}{]}. They integrated data including pitstructure and room count estimates, the presence of diagnostic architectural features, surface ceramic tallies, surveyor assessments, and limited dendrochronological data in an empirical Bayesian framework to develop population estimates for each region resolved to periods ranging from 20 to 125 years in length {[}\protect\hyperlink{ref-Ortman2016}{30},\protect\hyperlink{ref-Schwindt2016}{31}{]}. The VEP estimates are the highest quality regional palaeodemographic reconstructions currently available in the UUSW, and are completely independent of the radiocarbon records for the region. They are therefore useful in exploring the impact of investigation bias in the radiocarbon record.

@@

\caption{The UUSW as defined in this study (region with black border). The white dashed area represents the UUSW as defined in {[}\protect\hyperlink{ref-Bocinsky2016}{23}{]}. VEP study areas are white boxes. CMV: Central Mesa Verde; NRG: Northern Rio Grande.}\label{fig:Figure1}
\end{figure}

@@

Here we compare two large collections of radiocarbon and tree-ring dates drawn from the UUSW. After harmonizing those datasets with one another and calibrating the radiocarbon dates, we calculate binned summed probability distributions across the UUSW, CMV, and NRG. The radiocarbon dates we analyse here are drawn from the National Science Foundation supported project, ``Populating a Radiocarbon Database for North America,'' which have been uploaded to the Canadian Archaeological Radiocarbon Database (CARD; \url{https://www.canadianarchaeology.ca}). Tree-ring data are those presented in {[}\protect\hyperlink{ref-Bocinsky2016}{23}{]}. Both datasets are available on the Digital Archaeological Record (see Data Accessibility).

@@

The UUSW was geographically defined by {[}\protect\hyperlink{ref-Bocinsky2016}{23}{]} as the region between 105--113°W and 32--38°N. Here, we refine that definition in two ways. First, because location data for our radiocarbon database are only resolved to the county level for most sites {[}\protect\hyperlink{ref-Robinson2019}{26}{]}, we define our study region to include all counties whose geographic centroids fall within the region from {[}\protect\hyperlink{ref-Bocinsky2016}{23}{]}. Second, because of the generally poor tree-ring date record in the Sonoran Desert, we exclude the Phoenix and Tucson basins by dropping Maricopa, Pinal, and Pima counties in Arizona. Figure \ref{fig:Figure1} shows the original UUSW from {[}\protect\hyperlink{ref-Bocinsky2016}{23}{]} and revised study area used in this analysis, as well as the two VEP study areas.

@@

\begin{verbatim}
#> [1] "Calibrating radiocarbon ages..."
#> 
#> [1] "Done."
\end{verbatim}

@@

We draw on the tree-ring date database presented in {[}\protect\hyperlink{ref-Bocinsky2016}{23}{]}. These data were compiled by Tim Kohler and Rebecca Higgins from many contributors, and the vast majority of dates were produced by the Laboratory of Tree Ring Research at the University of Arizona. The original dataset contained 32,863 cutting, near-cutting, and non-cutting dates from across the US Southwest. Here, we only use the dates that are most likely to be accurate --- cutting dates (`B', `G', `L', `c', or `r' dates, with or without a `+') for which we know the year the tree died, and near-cutting dates (`v' or `v+' dates) which are likely within 0--3 years of the true date of the outermost ring. The {[}\protect\hyperlink{ref-Bocinsky2016}{23}{]} data contain site locations; for this analysis, we first spatially cropped the database to our study area, and then translated those locations to the county level. We converted the original date determinations to BP to match calibrated radiocarbon year conventions, and retained dates within 1750--150 BP.

@@

To prepare the radiocarbon dataset, we used functions provided by the rcarbon package for R {[}\protect\hyperlink{ref-Bevan2019}{1}{]}. We calibrated the radiocarbon dates using the methods of {[}\protect\hyperlink{ref-Bronk2008}{32}{]}, using the `intcal13' calibration curve {[}\protect\hyperlink{ref-Reimer2013}{33}{]}. To control for the inconsistent number of dates among the archaeological sites represented in the database (a form of investigation bias), we aggregated radiocarbon dates known to be from the same phase of an archaeological site using the rcarbon::binPrep function using the single linkage agglomeration method with an h-value of 100 (see {[}Supplementary Material{]} for a discussion of various h-values). This creates clusters of dates from the same site whose mean values (uncalibrated) are within 100 years of each other, and splits site occupations into distinct phases separated by 100 years or more. We constructed summed probability distributions (SPDs) from the calibrated dates using the \texttt{rcarbon::spd} function. We averaged probability distributions of binned dates, and then constructed SPDs from the resulting site-phase probability distributions.

@@

Figure \ref{fig:Figure4} presents radiocarbon and tree-ring SPDs for the UUSW, CMV, and NRG regions, as well as VEP population reconstructions for the CMV and NRG regions. Across the UUSW, we see striking agreement in directionality and timing between the radiocarbon and tree-ring series (Figure \ref{fig:Figure4}A); both series trend upwards until the late thirteenth century, and then decline rapidly. The tree-ring series --- the light orange line is the raw SPD, and the dark line is smoothed --- shows more variation on annual to decadal timescales, and yet matches the broad trends of the radiocarbon SPD. Roughly 150-year periodic fluctuations in the tree-ring SPD, described as periods of ``exploration'' and ``exploitation'' by {[}\protect\hyperlink{ref-Bocinsky2016}{23}{]}, are not present in the radiocarbon SPD, the result of a much smaller sample size of site-phases and lower temporal resolution of radiocarbon dating relative to tree-ring dating.

@@

\caption{Radiocarbon and tree-ring density distributions in the UUSW (A), CMV (B), and NRG (C) regions, and VEP population estimates for the CMV and NRG. The dark orange tree-ring density curve smooths the raw tree-ring density (light orange) using a 21-year center-aligned Gaussian kernel with a 5-year standard deviation, following {[}\protect\hyperlink{ref-Bocinsky2016}{23}{]}. In Panel A, the radiocarbon density (y-axis) is scaled by a factor of 3 to facilitate comparison with the tree-ring density over the target time period. The radiocarbon and tree-ring densities (y-axis) in panels B and C are scaled such that the area under their densities is equal to the area under the VEP population estimates (i.e., the cumulative population).}\label{fig:Figure4}
\end{figure}

Comparisons between the CMV and NRG regional SPDs and the VEP population estimates are also revealing. In Figures \ref{fig:Figure4}B and \ref{fig:Figure4}C, the regional SPDs have been scaled such that the areas under their densities (the orange and green lines) are equal to the area under the VEP population estimates (the black lines). For both regions, the radiocarbon and tree-ring SPDs capture the broad directionality of the VEP population estimates. In the CMV region (Figure \ref{fig:Figure4}B), the radiocarbon SPD has substantial probability mass prior to the VEP reconstruction, which is likely due to sampling bias in the radiocarbon record that is skewed in this region towards older (pre-Pueblo) sites. The radiocarbon SPD also does not capture the Pueblo I (AD 750--900) population peak (which may be too short-lived to be captured by such a small sample of radiocarbon dates), and seems to begin declining by the early thirteenth century AD. The tree-ring SPDs better capture the variability in population, though include a decline in the mid AD 1100s that is not in the VEP population estimates. In the NRG region (Figure \ref{fig:Figure4}C), both SPDs rise until about AD 1250, then decline thereafter. The high peak in the tree-ring record at AD 1250, far in excess of the VEP population estimate, is particularly interesting --- it likely signals the net impact of the immigration of many thousands of people to the NRG region, and the resulting (and simultaneous) burst in construction. Such a sudden and synchronous demographic and architectural event is hard to capture in the relatively lower resolution and sparsely sampled radiocarbon record, which tends to be smeared across sub-centennial scales courtesy of the calibration process {[}\protect\hyperlink{ref-Baillie1991}{34}{]}.

These results yield key insights on the relative sensitivity of the radiocarbon record to capture long-term, coarse-grained demographic change across the Southwest. As stated in the introduction, our aim is to understand how well the radiocarbon record captures population declines across the Southwest, as we need to develop a measure of confidence in radiocarbon records from regions across and beyond the Southwest that lack other demographic proxies. Both the radiocarbon and tree-ring SPDs show declines post-AD 1300 across all regions. In both the CMV and NRG the decline in the radiocarbon SPD begins earlier than in the VEP population series. This may indicate that radiocarbon time series make the decline seem to begin earlier than it actually does, which could be caused by a `cliff' in the radiocarbon calibration curve from AD 1200--1300 that is surrounded by two plateaus at AD 1100--1200 and AD 1400 {[}\protect\hyperlink{ref-Reimer2013}{33}{]}.

Comparisons of the radiocarbon and dendrochronological record with the architecture/ceramic record show similarities and differences that reinforce the need for multiple-proxy studies of palaeodemography when moving between broader and finer spatial scales. While the multi-proxy population reconstruction for the CMV region was similar to the radiocarbon and dendrochronological records, it showed a different trend for the NRG region (lower panel, Figure \ref{fig:Figure4}). There, the population reconstruction peaks from AD 1300--1400, whereas the radiocarbon record declines, at first quickly and then slowly. These different trends between the VEP population reconstructions and dendrochronological and radiocarbon proxies reflect some investigation bias in both the radiocarbon and dendrochronological records at smaller, intra-regional scales, which calls for caution in the use of radiocarbon records to assess possible paleo-migration scenarios at finer spatial scales. This result reinforces other work that has called for the use of multiple proxies as cross-checks on radiocarbon time series {[}\protect\hyperlink{ref-Williams2012}{9}{]}.

The central aim in comparing radiocarbon to dendrochronological records here was to assess the potential impacts of investigation bias in radiocarbon records, caused by researchers not taking radiocarbon samples in favour of other types of chronometric techniques (e.g., diagnostic artefact counts, tree-ring dates). Specifically, we are interested in the potential impact of this kind of bias in the post AD 1300 decline of radiocarbon SPDs across and beyond the Southwest. Similarities in the decline of both radiocarbon and dendrochronological records across the UUSW suggest that, at this particular regional scale, radiocarbon records reflect long-term, coarse-grained trends in paleodemography. Indeed, not only do we see corresponding declines in the radiocarbon and dendrochronological records at this large scale, but these patterns are also consistent with previous work on demographic and social change in the American Southwest: the Coalescent Communities Database, a Southwest-wide database of room counts and occupation spans of habitation sites with 12 rooms or more, supports a ``population decline that began between AD 1300 and 1350 and continued to European contact (Hill et al.~2004:Figures 2,4)'' {[}\protect\hyperlink{ref-Hill2004}{24},\protect\hyperlink{ref-Phillips2018}{35}{]}.

@@

\hypertarget{refs}{}
\begin{cslreferences}
\leavevmode\hypertarget{ref-Bevan2019}{}%
1. Bevan A, Palmisano A, Woodbridge J, Fyfe R, Roberts CN, Shennan S. 2019 The changing face of the Mediterranean---Land cover, demography and environmental change: Introduction and overview. \emph{The Holocene} \textbf{29}, 703--707. (doi:\href{https://doi.org/10.1177/0959683619826688}{10.1177/0959683619826688})

>>>>

@@

\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newenvironment{cslreferences}%
  {\setlength{\parindent}{0pt}%
  \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces}%
  {\par}

@@

The spatial and temporal ubiquity of archaeological radiocarbon dates makes them an unprecedented proxy for human paleodemography. The past decade has witnessed a rapid increase in the use of archaeological radiocarbon time series to reconstruct prehistoric population growth and decline in various regions of the world (Bevan et al., \protect\hyperlink{ref-Bevan2019}{2019}; Crema et al., \protect\hyperlink{ref-Crema2016}{2016}; Goldberg et al., \protect\hyperlink{ref-Goldberg2016}{2016}; Palmisano et al., \protect\hyperlink{ref-Palmisano2019}{2019}; Peros et al., \protect\hyperlink{ref-Peros2010}{2010}; Russell and Steele, \protect\hyperlink{ref-Russell2009}{2009}; Shennan et al., \protect\hyperlink{ref-Shennan2013}{2013}; Wang et al., \protect\hyperlink{ref-Wang2014}{2014}; Williams, \protect\hyperlink{ref-Williams2012}{2012}). One of the first aggregations of radiocarbon dates for palaeodemographic analysis was by Berry (Berry, \protect\hyperlink{ref-Berry1982}{1982}) in his study of population growth, decline and migration across the American Southwest. Radiocarbon time series were used as a complement to earlier periods that lacked robust dendrochronological records. Berry (\protect\hyperlink{ref-Berry1982}{1982}: 120) noted that they were ``not even remotely related to population size'', but rather the ``relative probability of occupation through time'' and ``the major trends in these charts (i.e., peaks and troughs) should provide a reliable indication of the direction of change''. Later, Rick (\protect\hyperlink{ref-Rick1987}{1987}: 56) more formally outlined the use of radiocarbon time series as a method for paleodemography by developing an `inference chain' that detailed the various biases involved in their use and interpretation. Rick identified three biases: creation bias, preservation bias, and investigation bias (Rick, \protect\hyperlink{ref-Rick1987}{1987}). Over the past decade, as archaeologists have assembled and analysed large radiocarbon datasets from various regions of the world, there has been considerable focus on these biases, notably creation (Crombé and Robinson, \protect\hyperlink{ref-Crombe2014}{2014}; Freeman, Byers, et al., \protect\hyperlink{ref-Freeman2018}{2018}; Naudinot et al., \protect\hyperlink{ref-Naudinot2014}{2014}) and preservation bias (Bluhm and Surovell, \protect\hyperlink{ref-Bluhm2019}{2019}; Peros et al., \protect\hyperlink{ref-Peros2010}{2010}; Surovell and Brantingham, \protect\hyperlink{ref-Surovell2007}{2007}; Surovell et al., \protect\hyperlink{ref-Surovell2009}{2009}; Williams, \protect\hyperlink{ref-Williams2012}{2012}), as well as bias caused by the radiocarbon calibration process (Armit et al., \protect\hyperlink{ref-Armit2013}{2013}; Bamforth and Grund, \protect\hyperlink{ref-Bamforth2012}{2012}; Contreras and Meadows, \protect\hyperlink{ref-Contreras2014}{2014}; Michczynski and Michczynska, \protect\hyperlink{ref-Michczynski2006}{2006}; Timpson et al., \protect\hyperlink{ref-Timpson2014}{2014}; Williams, \protect\hyperlink{ref-Williams2012}{2012}). Of the three biases Rick identified, investigation bias has been the most understudied thus far. Investigation bias occurs when investigators gather more radiocarbon samples for particular periods of time or specific geographic regions, which leads to periods or regions with more or less radiocarbon representation than the average time period or region. The biased selection of radiocarbon samples would then appear to be relative increases or decreases in population when they are merely artefacts of research history and fieldwork practice. Investigation bias has been understudied because for many regions of the world radiocarbon data are the only currently available proxy evidence enabling reconstructions of paleodemography in continuous time series spanning millennia. Assessing the impact of investigation bias on radiocarbon-based reconstructions of paleodemography requires comparison with other paleodemographic proxies.

The American Southwest provides one the best available case studies for assessing the role of investigation bias in radiocarbon-based reconstructions of paleodemography. It also has one of the most well-documented cases of population growth and decline in the world (Bocinsky et al., \protect\hyperlink{ref-Bocinsky2016}{2016}; Freeman, Baggio, et al., \protect\hyperlink{ref-Freeman2018PNAS}{2018}; Hill et al., \protect\hyperlink{ref-Hill2004}{2004}; Kohler et al., \protect\hyperlink{ref-Kohler2010b}{2010}; Robinson et al., \protect\hyperlink{ref-Robinson2019}{2019}). For almost a century, researchers have collected multiple proxies, such as dendrochronological records, settlement size records, and ceramic seriation records to address the magnitude of this population growth and decline. However, a critical knowledge gap still exists for the exact scale of population decline across and beyond the Southwest (Boyer et al., \protect\hyperlink{ref-Boyer2010}{2010}; Cameron, \protect\hyperlink{ref-Cameron2010}{2010}). This gap is because the less intensively researched regions lack multiple proxies for robust reconstructions of paleodemography. In these regions, radiocarbon data are the only available proxy time series for reconstructing paleodemography. Over the past six years, we have collected all available radiocarbon data across and beyond the Southwest (Robinson et al., \protect\hyperlink{ref-Robinson2019}{2019}). Initial analyses of this new radiocarbon dataset have revealed extensive population decline from AD 1300--1600 across and beyond the Southwest (Freeman, Baggio, et al., \protect\hyperlink{ref-Freeman2018PNAS}{2018}; Robinson et al., \protect\hyperlink{ref-Robinson2019}{2019}). Therefore, the geographic extent of Late Prehistoric population decline could be much broader than previously expected.

The potential importance of this widespread population decline for research on Late Prehistoric archaeology of the American West requires comparison of the sensitivity of lower-resolution radiocarbon data against other higher-resolution palaeodemographic proxies. In the most well-documented regions (such as the northern American Southwest) we expect radiocarbon records to reflect the highest amount of investigation bias, as researchers have the option of selecting from a suite of other higher-resolution proxies, including dendrochronology and ceramic seriation. Investigation bias can create an artificial `edge effect' spuriously appearing as a decrease in population. Here we develop a new method for arraying dendrochronological records as time series for comparison against radiocarbon time series. We focus on the broader region of the Upland US Southwest (UUSW) for which there is a robust dendrochronological record (Bocinsky et al., \protect\hyperlink{ref-Bocinsky2016}{2016}).

We also focus on the more concentrated and intensively investigated Central Mesa Verde (CMV) and Northern Rio Grande (NRG) regions of the UUSW (Figure \ref{fig:Figure1}), for which multiproxy palaeodemographic reconstructions were produced by the Village Ecodynamics Project (VEP) (Ortman, \protect\hyperlink{ref-Ortman2016}{2016}; Schwindt et al., \protect\hyperlink{ref-Schwindt2016}{2016}). The VEP is an interdisciplinary, multi-method research initiative aimed at better understanding demography of ancestral Pueblo farmers and human-environment interaction across the UUSW. As part of the project, VEP researchers cataloged data on all known archaeological habitation sites within a 4,600-km2 area in the CMV (Schwindt et al., \protect\hyperlink{ref-Schwindt2016}{2016}) and a 6,900-km2 area in the NRG (Ortman, \protect\hyperlink{ref-Ortman2016}{2016}), to which most of the CMV population migrated during the thirteenth century AD (Ortman, \protect\hyperlink{ref-Ortman2016}{2016}). They integrated data including pitstructure and room count estimates, the presence of diagnostic architectural features, surface ceramic tallies, surveyor assessments, and limited dendrochronological data in an empirical Bayesian framework to develop population estimates for each region resolved to periods ranging from 20 to 125 years in length (Ortman, \protect\hyperlink{ref-Ortman2016}{2016}; Schwindt et al., \protect\hyperlink{ref-Schwindt2016}{2016}). The VEP estimates are the highest quality regional palaeodemographic reconstructions currently available in the UUSW, and are completely independent of the radiocarbon records for the region. They are therefore useful in exploring the impact of investigation bias in the radiocarbon record.

@@

\caption{The UUSW as defined in this study (region with black border). The white dashed area represents the UUSW as defined by Bocinsky et al. (\protect\hyperlink{ref-Bocinsky2016}{2016}). VEP study areas are white boxes. CMV: Central Mesa Verde; NRG: Northern Rio Grande.}\label{fig:Figure1}
\end{figure}

@@

Here we compare two large collections of radiocarbon and tree-ring dates drawn from the UUSW. After harmonizing those datasets with one another and calibrating the radiocarbon dates, we calculate binned summed probability distributions across the UUSW, CMV, and NRG. The radiocarbon dates we analyse here are drawn from the National Science Foundation supported project, ``Populating a Radiocarbon Database for North America,'' which have been uploaded to the Canadian Archaeological Radiocarbon Database (CARD; \url{https://www.canadianarchaeology.ca}). Tree-ring data are those presented by Bocinsky et al. (\protect\hyperlink{ref-Bocinsky2016}{2016}). Both datasets are available on the Digital Archaeological Record (see \protect\hyperlink{data-accessibility}{Data Accessibility}).

@@

The UUSW was geographically defined by Bocinsky et al. (\protect\hyperlink{ref-Bocinsky2016}{2016}) as the region between 105--113°W and 32--38°N. Here, we refine that definition in two ways. First, because location data for our radiocarbon database are only resolved to the county level for most sites (Robinson et al., \protect\hyperlink{ref-Robinson2019}{2019}), we define our study region to include all counties whose geographic centroids fall within the region from Bocinsky et al. (\protect\hyperlink{ref-Bocinsky2016}{2016}). Second, because of the generally poor tree-ring date record in the Sonoran Desert, we exclude the Phoenix and Tucson basins by dropping Maricopa, Pinal, and Pima counties in Arizona. Figure \ref{fig:Figure1} shows the original UUSW from Bocinsky et al. (\protect\hyperlink{ref-Bocinsky2016}{2016}) and revised study area used in this analysis, as well as the two VEP study areas.

@@

@@

We draw on the tree-ring date database presented in (Bocinsky et al., \protect\hyperlink{ref-Bocinsky2016}{2016}). These data were compiled by Tim Kohler and Rebecca Higgins from many contributors, and the vast majority of dates were produced by the Laboratory of Tree Ring Research at the University of Arizona. The original dataset contained 32,863 cutting, near-cutting, and non-cutting dates from across the US Southwest. Here, we only use the dates that are most likely to be accurate --- cutting dates (`B', `G', `L', `c', or `r' dates, with or without a `+') for which we know the year the tree died, and near-cutting dates (`v' or `v+' dates) which are likely within 0--3 years of the true date of the outermost ring. The (Bocinsky et al., \protect\hyperlink{ref-Bocinsky2016}{2016}) data contain site locations; for this analysis, we first spatially cropped the database to our study area, and then translated those locations to the county level. We converted the original date determinations to BP to match calibrated radiocarbon year conventions, and retained dates within 1750--150 BP.

@@

To prepare the radiocarbon dataset, we used functions provided by the rcarbon package for R (Bevan et al., \protect\hyperlink{ref-Bevan2019}{2019}). We calibrated the radiocarbon dates using the methods of Bronk Ramsey (\protect\hyperlink{ref-Bronk2008}{2008}), using the `intcal13' calibration curve (Reimer et al., \protect\hyperlink{ref-Reimer2013}{2013}). To control for the inconsistent number of dates among the archaeological sites represented in the database (a form of investigation bias), we aggregated radiocarbon dates known to be from the same phase of an archaeological site using the rcarbon::binPrep function using the single linkage agglomeration method with an h-value of 100 (see {[}Supplementary Material{]} for a discussion of various h-values). This creates clusters of dates from the same site whose mean values (uncalibrated) are within 100 years of each other, and splits site occupations into distinct phases separated by 100 years or more. We constructed summed probability distributions (SPDs) from the calibrated dates using the \texttt{rcarbon::spd} function. We averaged probability distributions of binned dates, and then constructed SPDs from the resulting site-phase probability distributions.\footnote{Using this method, a site-phase with a single date and a site-phase with 10 dates would each contribute equally to the resulting SPD. This method of developing the SPD is therefore a model of the relative abundance of site occupation events through time rather than of dates through time. This method does not account for investigation biases deriving from the definition of ``site'' or the decision to use a particular dating technique at a site in the first place.}

@@

Figure \ref{fig:Figure4} presents radiocarbon and tree-ring SPDs for the UUSW, CMV, and NRG regions, as well as VEP population reconstructions for the CMV and NRG regions. Across the UUSW, we see striking agreement in directionality and timing between the radiocarbon and tree-ring series (Figure \ref{fig:Figure4}A); both series trend upwards until the late thirteenth century, and then decline rapidly. The tree-ring series --- the light orange line is the raw SPD, and the dark line is smoothed --- shows more variation on annual to decadal timescales, and yet matches the broad trends of the radiocarbon SPD. Roughly 150-year periodic fluctuations in the tree-ring SPD, described as periods of ``exploration'' and ``exploitation'' by Bocinsky et al. (\protect\hyperlink{ref-Bocinsky2016}{2016}), are not present in the radiocarbon SPD, the result of a much smaller sample size of site-phases and lower temporal resolution of radiocarbon dating relative to tree-ring dating.

@@

\caption{Radiocarbon and tree-ring density distributions in the UUSW (A), CMV (B), and NRG (C) regions, and VEP population estimates for the CMV and NRG. The dark orange tree-ring density curve smooths the raw tree-ring density (light orange) using a 21-year center-aligned Gaussian kernel with a 5-year standard deviation, following Bocinsky et al. (\protect\hyperlink{ref-Bocinsky2016}{2016}). In Panel A, the radiocarbon density (y-axis) is scaled by a factor of 3 to facilitate comparison with the tree-ring density over the target time period. The radiocarbon and tree-ring densities (y-axis) in panels B and C are scaled such that the area under their densities is equal to the area under the VEP population estimates (i.e., the cumulative population).}\label{fig:Figure4}
\end{figure}

Comparisons between the CMV and NRG regional SPDs and the VEP population estimates are also revealing. In Figures \ref{fig:Figure4}B and \ref{fig:Figure4}C, the regional SPDs have been scaled such that the areas under their densities (the orange and green lines) are equal to the area under the VEP population estimates (the black lines). For both regions, the radiocarbon and tree-ring SPDs capture the broad directionality of the VEP population estimates. In the CMV region (Figure \ref{fig:Figure4}B), the radiocarbon SPD has substantial probability mass prior to the VEP reconstruction, which is likely due to sampling bias in the radiocarbon record that is skewed in this region towards older (pre-Pueblo) sites. The radiocarbon SPD also does not capture the Pueblo I (AD 750--900) population peak (which may be too short-lived to be captured by such a small sample of radiocarbon dates), and seems to begin declining by the early thirteenth century AD. The tree-ring SPDs better capture the variability in population, though include a decline in the mid AD 1100s that is not in the VEP population estimates. In the NRG region (Figure \ref{fig:Figure4}C), both SPDs rise until about AD 1250, then decline thereafter. The high peak in the tree-ring record at AD 1250, far in excess of the VEP population estimate, is particularly interesting --- it likely signals the net impact of the immigration of many thousands of people to the NRG region, and the resulting (and simultaneous) burst in construction. Such a sudden and synchronous demographic and architectural event is hard to capture in the relatively lower resolution and sparsely sampled radiocarbon record, which tends to be smeared across sub-centennial scales courtesy of the calibration process (Baillie, \protect\hyperlink{ref-Baillie1991}{1991}).

These results yield key insights on the relative sensitivity of the radiocarbon record to capture long-term, coarse-grained demographic change across the Southwest. As stated in the introduction, our aim is to understand how well the radiocarbon record captures population declines across the Southwest, as we need to develop a measure of confidence in radiocarbon records from regions across and beyond the Southwest that lack other demographic proxies. Both the radiocarbon and tree-ring SPDs show declines post-AD 1300 across all regions. In both the CMV and NRG the decline in the radiocarbon SPD begins earlier than in the VEP population series. This may indicate that radiocarbon time series make the decline seem to begin earlier than it actually does, which could be caused by a `cliff' in the radiocarbon calibration curve from AD 1200--1300 that is surrounded by two plateaus at AD 1100--1200 and AD 1400 (Reimer et al., \protect\hyperlink{ref-Reimer2013}{2013}).

Comparisons of the radiocarbon and dendrochronological record with the architecture/ceramic record show similarities and differences that reinforce the need for multiple-proxy studies of palaeodemography when moving between broader and finer spatial scales. While the multi-proxy population reconstruction for the CMV region was similar to the radiocarbon and dendrochronological records, it showed a different trend for the NRG region (lower panel, Figure \ref{fig:Figure4}). There, the population reconstruction peaks from AD 1300--1400, whereas the radiocarbon record declines, at first quickly and then slowly. These different trends between the VEP population reconstructions and dendrochronological and radiocarbon proxies reflect some investigation bias in both the radiocarbon and dendrochronological records at smaller, intra-regional scales, which calls for caution in the use of radiocarbon records to assess possible paleo-migration scenarios at finer spatial scales. This result reinforces other work that has called for the use of multiple proxies as cross-checks on radiocarbon time series (Williams, \protect\hyperlink{ref-Williams2012}{2012}, p. @Crombe2014).

The central aim in comparing radiocarbon to dendrochronological records here was to assess the potential impacts of investigation bias in radiocarbon records, caused by researchers not taking radiocarbon samples in favour of other types of chronometric techniques (e.g., diagnostic artefact counts, tree-ring dates). Specifically, we are interested in the potential impact of this kind of bias in the post AD 1300 decline of radiocarbon SPDs across and beyond the Southwest. Similarities in the decline of both radiocarbon and dendrochronological records across the UUSW suggest that, at this particular regional scale, radiocarbon records reflect long-term, coarse-grained trends in paleodemography. Indeed, not only do we see corresponding declines in the radiocarbon and dendrochronological records at this large scale, but these patterns are also consistent with previous work on demographic and social change in the American Southwest: the Coalescent Communities Database, a Southwest-wide database of room counts and occupation spans of habitation sites with 12 rooms or more, supports a ``population decline that began between AD 1300 and 1350 and continued to European contact'' (Phillips et al., \protect\hyperlink{ref-Phillips2018}{2018}, following @Hill2004: Figures 2,4).

@@

\hypertarget{refs}{}
\begin{cslreferences}
\leavevmode\hypertarget{ref-Armit2013}{}%
Armit, I., Swindles, G.T., Becker, K., 2013. From dates to demography in later prehistoric Ireland? Experimental approaches to the meta-analysis of large \textsuperscript{14}C data-sets. Journal of archaeological science 40, 433--438. \url{https://doi.org/10.1016/j.jas.2012.08.039}

\leavevmode\hypertarget{ref-Baillie1991}{}%
Baillie, M.G., 1991. Suck-in and smear: Two related chronological problems for the 90s. Journal of theoretical archaeology 2, 12--16.

\leavevmode\hypertarget{ref-Bamforth2012}{}%
Bamforth, D.B., Grund, B., 2012. Radiocarbon calibration curves, summed probability distributions, and early Paleoindian population trends in North America. Journal of Archaeological Science 39, 1768--1774. \url{https://doi.org/10.1016/j.jas.2012.01.017}

\leavevmode\hypertarget{ref-Berry1982}{}%
Berry, M.S., 1982. Time, space and transition in anasazi prehistory. University of Utah Press, Salt Lake City, Utah.
